---
title: "SVM and ANN"
author: "Dibanisa Fakude"
date: "2024-05-12"
output: pdf_document
---

```{r}
library(tidyverse)
library(caTools)
df<- read.csv("heart_failure_clinical_records_dataset.csv")
```
Data wrangling
```{r}
#Converting the factors variables
df<-df %>% 
  mutate(anaemia = factor(anaemia,levels= c(0,1),labels = c("No","Yes")),
         diabetes= factor(diabetes,levels= c(0,1),labels = c("No","Yes")),
         high_blood_pressure= factor(high_blood_pressure,levels= c(0,1),labels = c("No","Yes")),                   sex = factor(sex,levels= c(0,1),labels = c("F","M")),
         smoking= factor(smoking,levels= c(0,1),labels = c("No","Yes")),
         DEATH_EVENT = factor(DEATH_EVENT,levels= c(0,1),labels = c("No","Yes")))
# Distribution of categorical variables
cat_vars <- select_if(df, is.factor)
table(cat_vars)

# Distribution of numerical variables
num_vars <- select_if(df, is.numeric)
summary(num_vars)

#Correlation
correlation<-cor(num_vars)
```
Data exploration

```{r}
#Death event data count , checking imbalances
ggplot(df, aes(x = DEATH_EVENT)) +
  geom_bar(fill = "grey",color = "black") +
  labs(title = "Distribution of DEATH_EVENT", x = "DEATH_EVENT", y = "Count")+
  theme_bw()+                         
  theme(plot.title = element_text(hjust = 0.5))

```
```{r}
ggplot(df, aes(x = sex, fill = factor(sex))) +
  geom_bar(color = "black") +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count") +
  theme_bw() +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"), name = "Sex")+                         
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
# Load necessary packages
library(ggplot2)

# Create the stacked bar plot
ggplot(df, aes(x = sex, fill = factor(DEATH_EVENT))) +
  geom_bar(color = "black", position = "stack") +
  labs(title = "Death Event Distribution by Sex", x = "Sex", y = "Count", fill = "Death Event") +
  theme_bw() +
  scale_fill_manual(values = c("#E69F00", "#56B4E9"), 
                    labels = c("No Death Event", "Death Event"))+                         
  theme(plot.title = element_text(hjust = 0.5))

```
```{r}
# Plotting correlation heatmap with 
corrplot(correlation, method = "color",
         col = colorRampPalette(c("yellow", "red"))(10),  
         addCoef.col = "black", number.cex = 0.6,      
         tl.cex = 0.5)                                    


```


a) Randomly split the data into training and testing sets by applying an 80-20 split.
```{r}
#(a)
# Split data into training and testing sets
set.seed(123)  # For reproducibility
split <- sample.split(df$DEATH_EVENT, SplitRatio = 0.8)
x_train <- subset(df, split == TRUE)
x_test <- subset(df, split == FALSE)

# Perform upsampling on the training set
oversampled_x_train <- upSample(x = x_train[ , -ncol(x_train)], 
                                y = x_train$DEATH_EVENT, 
                                list = FALSE)

# Combine the upsampled features and the target variable into a data frame
oversampled_x_train <- cbind(oversampled_x_train, DEATH_EVENT = oversampled_x_train$Class)
oversampled_x_train$Class <- NULL

# Verify the class distribution in the upsampled training set
table(oversampled_x_train$DEATH_EVENT)
```

b) Build a support vector machine using a radial kernel function with cost=0.1 and
gamma=0.1. Report the classification accuracy, recall, specificity, F1 score, and
ROC AUC.

```{r}
#(b)

library(e1071)
library(pROC)
set.seed(123)

# Train SVM classifier
classifier <- svm(DEATH_EVENT ~ .,
                  data = oversampled_x_train,
                  scale = TRUE,
                  type = "C-classification",
                  kernel = "radial",
                  cost = 0.1,
                  gamma = 0.1)

# Make predictions on test dataset
y_pred <- predict(classifier, newdata = x_test[-13])

# Evaluate classifier performance with confusion matrix
cm <- table(y_pred, x_test[, 13])


true_positives <- cm[2, 2]
false_positives <- cm[1, 2]
true_negatives <- cm[1, 1]
false_negatives <- cm[2, 1]

# Calculate metrics
accuracy <- (true_positives + true_negatives) / sum(cm)
recall <- true_positives / (true_positives + false_negatives)
specificity <- true_negatives / (true_negatives + false_positives)
precision <- true_positives / (true_positives + false_positives)
F1_score <- 2 * (precision * recall) / (precision + recall)
roc_response <- roc(x_test[, 13], as.numeric(y_pred))
auc_value <- auc(roc_response)
# Print the metrics
cat("Classification Accuracy:", accuracy, "\n")
cat("Recall:", recall, "\n")
cat("Specificity:", specificity, "\n")
cat("F1 Score:", F1_score, "\n")
cat("ROC AUC:", auc_value, "\n")

```

c) Repeat steps (a) and (b) 100 times and report the average of each of these metrics
over the 100 runs. Also, provide a boxplot for each metric’s 100 values.
```{r}
library(tidyverse)
library(caTools)
library(e1071)
library(pROC)

# Initialize empty vectors to store metric values
accuracy_vec <- vector("numeric", length = 100)
recall_vec <- vector("numeric", length = 100)
specificity_vec <- vector("numeric", length = 100)
F1_score_vec <- vector("numeric", length = 100)
auc_value_vec <- vector("numeric", length = 100)

# Repeat the process 100 times
for (i in 1:100) {
  set.seed(123 + i)  # Set seed for reproducibility
  
  # Split the data
  split <- sample.split(df$DEATH_EVENT, SplitRatio = 0.8)
  x_train <- subset(df, split == TRUE)
  x_test <- subset(df, split == FALSE)
  
  # Train SVM classifier
  classifier <- svm(DEATH_EVENT ~ .,
                    data = oversampled_x_train,
                    scale = TRUE,
                    type = "C-classification",
                    kernel = "radial",
                    cost = 0.1,
                    gamma = 0.1)
  
  # Make predictions on test dataset
  y_pred <- predict(classifier, newdata = x_test[-13])
  
  # Evaluate classifier performance with confusion matrix
  cm <- table(y_pred, x_test[, 13])
  
  # Extract values from confusion matrix
  true_positives <- cm[2, 2]
  false_positives <- cm[1, 2]
  true_negatives <- cm[1, 1]
  false_negatives <- cm[2, 1]
  
  # Calculate metrics
  accuracy_vec[i] <- (true_positives + true_negatives) / sum(cm)
  recall_vec[i] <- true_positives / (true_positives + false_negatives)
  specificity_vec[i] <- true_negatives / (true_negatives + false_positives)
  precision <- true_positives / (true_positives + false_positives)
  F1_score_vec[i] <- 2 * (precision * recall_vec[i]) / (precision + recall_vec[i])
  roc_response <- roc(x_test[, 13], as.numeric(y_pred))
  auc_value_vec[i] <- auc(roc_response)
}

# Calculate averages of metrics
avg_accuracy <- mean(accuracy_vec)
avg_recall <- mean(recall_vec)
avg_specificity <- mean(specificity_vec)
avg_F1_score <- mean(F1_score_vec)
avg_auc_value <- mean(auc_value_vec)

# Create a dataframe to store metric values
metrics_df <- data.frame(
  Accuracy = accuracy_vec,
  Recall = recall_vec,
  Specificity = specificity_vec,
  F1_Score = F1_score_vec,
  AUC = auc_value_vec
)

# Create boxplots for each metric
boxplot(metrics_df, main = "Performance Metrics Distribution", col = "grey",
        ylab = "Metric Value")

# Print average values of each metric
cat("Average Accuracy:", avg_accuracy, "\n")
cat("Average Recall:", avg_recall, "\n")
cat("Average Specificity:", avg_specificity, "\n")
cat("Average F1 Score:", avg_F1_score, "\n")
cat("Average AUC Value:", avg_auc_value, "\n")
```
d) Repeat steps (a), (b) and (c) with different cost and gamma parameters. You are
expected to explore at least 3 different values for each hyperparameter. Report
your findings using appropriate tables and/or graphs.
```{r}
set.seed(123)
# Define different values for cost and gamma
cost_values <- c(0.1, 1, 10)
gamma_values <- c(0.1, 1, 10)

# Initialize empty data frames to store results
results <- data.frame(Cost = numeric(), Gamma = numeric(), Accuracy = numeric(),
                      Recall = numeric(), Specificity = numeric(),
                      F1_Score = numeric(), AUC = numeric())

# Iterate over all combinations of cost and gamma values
for (cost in cost_values) {
  for (gamma in gamma_values) {
    set.seed(123)  # Set seed for reproducibility
    
    # Train SVM classifier with current cost and gamma
    classifier <- svm(DEATH_EVENT ~ .,
                      data = oversampled_x_train,
                      scale = TRUE,
                      type = "C-classification",
                      kernel = "radial",
                      cost = cost,
                      gamma = gamma)
    
    # Make predictions on test dataset
    y_pred <- predict(classifier, newdata = x_test[-13])
    
    # Evaluate classifier performance with confusion matrix
    cm <- table(y_pred, x_test[, 13])
    
    # Extract values from confusion matrix
    true_positives <- cm[2, 2]
    false_positives <- cm[1, 2]
    true_negatives <- cm[1, 1]
    false_negatives <- cm[2, 1]
    
    # Calculate metrics
    accuracy <- (true_positives + true_negatives) / sum(cm)
    recall <- true_positives / (true_positives + false_negatives)
    specificity <- true_negatives / (true_negatives + false_positives)
    precision <- true_positives / (true_positives + false_positives)
    F1_score <- 2 * (precision * recall) / (precision + recall)
    roc_response <- roc(x_test[, 13], as.numeric(y_pred))
    auc_value <- auc(roc_response)
    
    # Store results in data frame
    results <- rbind(results, data.frame(Cost = cost, Gamma = gamma,
                                         Accuracy = accuracy, Recall = recall,
                                         Specificity = specificity,
                                         F1_Score = F1_score, AUC = auc_value))
  }
}

# Print results
print(results)


# Reshape data for plotting
library(reshape2)
results_melted <- melt(results, id.vars = c("Cost", "Gamma"))

# Plot heatmap for average accuracy
ggplot(results_melted, aes(x = Cost, y = Gamma, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "red", high = "green") +
  facet_wrap(~variable, scales = "free", ncol = 2) +
  labs(title = "Performance Metrics Heatmap", x = "Cost", y = "Gamma")
  labs(x = "Cost", y = "Gamma", fill = "Accuracy") +
  theme_bw() 



```
e) Tune the cost and gamma parameters for 2 using grid search on the training
data from (a) and report your best model parameter. Test your best model
performance on the test data from (a).

```{r}
set.seed(123)
library(e1071)

# Define a grid of cost and gamma values to search over
cost_grid <- c(0.01, 0.1, 1, 10, 100)
gamma_grid <- c(0.01, 0.1, 1, 10, 100)

# Initialize variables to store best parameters and performance
best_accuracy <- 0
best_cost <- NULL
best_gamma <- NULL

# Perform grid search
for (cost in cost_grid) {
  for (gamma in gamma_grid) {
    # Train SVM classifier with current cost and gamma
    classifier <- svm(DEATH_EVENT ~ .,
                      data = oversampled_x_train,
                      scale = TRUE,
                      type = "C-classification",
                      kernel = "radial",
                      cost = cost,
                      gamma = gamma)
    
    # Make predictions on training data
    y_pred_train <- predict(classifier, newdata = oversampled_x_train[-13])
    
    # Calculate accuracy on training data
    accuracy <- sum(y_pred_train == oversampled_x_train$DEATH_EVENT) / nrow(oversampled_x_train)
    
    # Update best parameters if current model is better
    if (accuracy > best_accuracy) {
      best_accuracy <- accuracy
      best_cost <- cost
      best_gamma <- gamma
    }
  }
}

# Train SVM classifier with best parameters
best_classifier <- svm(DEATH_EVENT ~ .,
                       data = oversampled_x_train,
                       scale = TRUE,
                       type = "C-classification",
                       kernel = "radial",
                       cost = best_cost,
                       gamma = best_gamma)

# Make predictions on test data
y_pred_test <- predict(best_classifier, newdata = x_test[-13])

# Calculate accuracy on test data
test_accuracy <- sum(y_pred_test == x_test$DEATH_EVENT) / nrow(x_test)

# Report best model parameters and test performance
cat("Best model parameters:\n")
cat("Cost:", best_cost, "\n")
cat("Gamma:", best_gamma, "\n")
cat("Test Accuracy:", test_accuracy, "\n")

```



